---
title: "Global Baseline Estimate"
author: "Johnny Maldonado"
date: "2026-09-16"
output: html_document
---

## Introduction

The goal of this project is to use the survey data I collected on movie ratings to develop a non-personalized recommendation using the Global Baseline Estimate method. I plan to first examine the structure and distribution of the ratings, then calculate the overall average rating and compare individual user and movie ratings against that average to determine the adjustments needed for the Global Baseline Estimate. I will also review the dataset for missing ratings, inconsistent formatting, or other issues that may need to be addressed before completing the recommendation analysis.

### Load Packages

```{r loading libraries}

library(dplyr)
library(readr)
library(ggplot2)
library(tidyr)
```
### Load data
```{r load data}
movie_ratings <-read_csv("https://github.com/jmald1987/DATA607-Movie_Ratings/raw/refs/heads/main/movie_ratings.csv"
)
```

### Examine the Data

Before calculating the Global Baseline Estimate, I examined the structure and summary of the dataset to identify the variables, data types, and missing ratings.

```{r examine-data}
glimpse(movie_ratings)
summary(movie_ratings)
```

The dataset contains 77 observations and four variables: `name`, `title`, `media_type`, and `rating`. The initial examination also shows that `rating` was imported as a character variable because unrated titles are represented by the text `"NULL"`.

### Prepare the Data

Before calculating averages, the `"NULL"` values need to be treated as missing values and the remaining ratings converted from character values to numeric values.

```{r prepare-data}
movie_ratings <- movie_ratings %>%
  mutate(
    rating = na_if(rating, "NULL"),
    rating = as.numeric(rating)
  )

glimpse(movie_ratings)
summary(movie_ratings)
```

After cleaning, `rating` is numeric and unrated movies are represented as `NA`. These missing ratings can now be excluded from the average calculations while remaining available as the ratings that may later be estimated.

### Calculate the Global Mean

The global mean represents the average of all available ratings in the dataset and serves as the starting point for the Global Baseline Estimate.

```{r global-mean}
global_mean <- mean(movie_ratings$rating, na.rm = TRUE)

global_mean
```

### Calculate User Averages and Effects

Next, the average rating for each user is calculated. The user effect measures the difference between each user's average rating and the global mean.

```{r user-effects}
user_effects <- movie_ratings %>%
  group_by(name) %>%
  summarise(
    user_average = mean(rating, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    user_effect = user_average - global_mean
  )

user_effects
```

A positive user effect indicates that the user tends to give ratings above the overall average, while a negative effect indicates that the user tends to give lower ratings.

### Calculate Movie Averages and Effects

The same process is used for each title. The movie effect measures the difference between a movie's average rating and the global mean.

```{r movie-effects}
movie_effects <- movie_ratings %>%
  group_by(title) %>%
  summarise(
    movie_average = mean(rating, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    movie_effect = movie_average - global_mean
  )

movie_effects
```

A positive movie effect indicates that the title was generally rated above the global average, while a negative value indicates that it was generally rated below the global average.

### Calculate the Global Baseline Estimates

The Global Baseline Estimate combines the global mean with the user and movie effects. The calculation is applied to the missing ratings to estimate how each user might rate a title they did not rate in the original survey.

```{r baseline-estimates}
baseline_estimates <- movie_ratings %>%
  filter(is.na(rating)) %>%
  select(name, title, media_type) %>%
  left_join(user_effects, by = "name") %>%
  left_join(movie_effects, by = "title") %>%
  mutate(
    estimated_rating = global_mean + user_effect + movie_effect
  )

baseline_estimates
```

### Identify Recommendations

The estimated ratings can be sorted from highest to lowest to identify the strongest potential recommendations for each user.

```{r recommendations, corrected grouping} 
recommendations <- baseline_estimates %>%
  group_by(name) %>%
  slice_max(
    order_by = estimated_rating,
    n = 1,
    with_ties = FALSE
  ) %>%
  ungroup() %>%
  select(name, title, media_type, estimated_rating)

recommendations
```

## Conclusions 


The Global Baseline Estimate uses the overall rating average along with user and title effects to estimate missing ratings. These estimates are then used to identify the highest predicted title for each user. This approach accounts for differences in how individual users rate titles as well as differences in how individual titles are rated overall.
